Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-10 wpe.tex
more updates

Here we will be utilizing a variation of the permutation entropy, the
\emph{weighted permutation entropy} (WPE)~\cite{fadlallah2013}. The weighted
permutation entropy attempts to correct for observational noise which is larger
than some trends in the data, but smaller than the larger scale features --- for
example, a signal that switches between two fixed points with noise about those
fixed points. The weighted permutation entropy would be dominated by the
switching rather than by the stochastic fluctuation. To accomplish this, the
\emph{weight} of a permutation is taken into account:
\begin{align*}
  w(x_{t+1:t+n}) = \frac{1}{n}
                 \sum_{x_i \in \{x_{t+1:t+n}\}}
                 \left( x_i - \bar{x}_{t+1:t+n} \right)^2
\end{align*}
where $x_{t+1:t+n}$ is a sequence of values $x_{t+1}, \ldots, x_{t+n}$, and
$\bar{x}_{t+1:t+n}$ is the arithmetic mean of those values.

The weighted probability of a permutation is then:
\begin{align*}
  p_w(\pi) = \frac{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n}) \cdot \delta(\phi(x_{t:t+n}), \pi) }{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n})}
\end{align*}
where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively, this weighted
probability enhances permutations involved in ``large'' features and demotes
permutations which are small in amplitude relative to the features of the time
series. The weighted permutation entropy is then:
\begin{align*}
  H_w(n) = - \sum_{\pi \in \mathcal{S}_n} p_w(\pi) \log_2 p_w(\pi),
\end{align*}
which can also be normalized by dividing by $\log_2(n!)$, and will be in all the
results of this paper.

In practice, calculating permutation entropy and weighted permutation entropy
involves choosing a good value for the word length $n$. The primary
consideration is that the value be large enough that forbidden ordinals are
discovered, yet small enough that reasonable statistics over the ordinals are
gathered: e.g.:
\begin{align*}
  n = \operatornamewithlimits{argmax}_\ell \{ T \gtrapprox 100 \ell! \},
\end{align*}
assuming an average of 100 counts per ordinal is sufficient. In the literature,
$3 \le n \le 6$ is a standard choice --- generally without any formal
justification. In theory, the permutation entropy should reach an asymptote with
increasing $n$, but that requires an arbitrarily long time series. In practice,
what one should do is calculate the \emph{persistent} permutation entropy by
increasing $n$ until the result converges, but data length issues can intrude
before that convergence is reached.

>>>>

Here we will be utilizing a variation of the permutation entropy, the \emph{weighted permutation entropy} (WPE)~\cite{fadlallah2013}. The weighted permutation entropy attempts to correct for observational noise which is larger than some trends in the data, but smaller than the larger scale features --- for example, a signal that switches between two fixed points with noise about those fixed points. The weighted permutation entropy would be dominated by the switching rather than by the stochastic fluctuation. To accomplish this, the \emph{weight} of a permutation is taken into account:
\begin{align*}
  w(x_{t+1:t+\ell}) = \frac{1}{\ell} \sum_{x_i \in \{x_{t+1:t+\ell}\}}
                      \left( x_i - \bar{x}_{t+1:t+\ell} \right)^2
\end{align*}
where $x_{t+1:t+\ell}$ is a sequence of values $x_{t+1}, \ldots, x_{t+\ell}$, and $\bar{x}_{t+1:t+\ell}$ is the arithmetic mean of those values.

The weighted probability of a permutation is then:
\begin{align*}
  p_w(\pi) = \frac{\displaystyle \sum_{t \le T - \ell} w(x_{t+1:t+\ell}) \cdot \delta(\phi(x_{t:t+\ell}), \pi) }{\displaystyle \sum_{t \le T - \ell} w(x_{t+1:t+\ell})}
\end{align*}
where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively, this weighted probability enhances permutations involved in ``large'' features and demotes permutations which are small in amplitude relative to the features of the time series. The weighted permutation entropy is then:
\begin{align*}
  H_w(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} p_w(\pi) \log_2 p_w(\pi),
\end{align*}
which can also be normalized by dividing by $\log_2(\ell!)$, and will be in all the results of this paper.

In practice, calculating permutation entropy and weighted permutation entropy involves choosing a good value for the word length $\ell$. The primary consideration is that the value be large enough that forbidden ordinals are discovered, yet small enough that reasonable statistics over the ordinals are gathered: e.g.:
\begin{align*}
  \ell = \operatornamewithlimits{argmax}_n \{ T \gtrapprox 100 n! \},
\end{align*}
assuming an average of 100 counts per ordinal is sufficient. In the literature, $3 \le \ell \le 6$ is a standard choice --- generally without any formal justification. In theory, the permutation entropy should reach an asymptote with increasing $\ell$, but that requires an arbitrarily long time series. In practice, what one should do is calculate the \emph{persistent} permutation entropy by increasing $\ell$ until the result converges, but data length issues can intrude before that convergence is reached. Here, we found that $\ell = 6$ was a good choice, because with $\ell \geq 7$ many signals had entropies plummet due to finite data effects and so with a word length of 6 we are capturing as much structure as possible without sacraficing the accuracy of ordinal statistics.
